---
title: "SWMPr Workshop Part 1: Refresher on reading in data, and beginning to plot"
author: "Kim Cressman"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

<br>

#### Step 1: Set the Working Directory


1.  From the Rstudio menu bar, select 'Session' --> 'Set Working Directory' --> 'Choose Directory' (or use keyboard shortcut Ctrl + Shift + H) 
2.  You will see in the console a line that looks like this:

```{r, opts_chunk$set(message=FALSE, warning=FALSE)}
setwd("C:/Users/kimberly.cressman/Desktop/NERRS SWMPr course 2016")
```
  +  notice the forward slashes!
      +  you may have to change yours if you're copying from windows explorer
  +  and the quotation marks!
  +  and the capitalization.
      +  R is sensitive to capitalization.


3. Copy and paste that into your script window so you can just default to it every time you use this script
    +  if Rstudio didn't open with a blank script, click on the plus sign to create one.


***

#### Step 2: Read in a file (from your working directory)

Use the read.csv() command.

*  The file name needs to be inside quotation marks (double "" or single '')
*  Make sure the capitalization is correct
*  Remember to include the file extension (.csv)
*  Make sure your file is a .csv (there are packages that will read in Excel files for you, but .csv is simplist)
    + To convert a .xlsx file to a .csv, go to File --> Save as --> CSV (Comma Delimited)

<br>
`read.csv("GNDPCWQ2015.csv")`

<br>
To run a line of code:

*  from the console: Enter
*  from a script - __you should use scripts__
    +  Ctrl + Enter   OR
    +  Ctrl + R   OR
    +  Run button in upper right corner of script panel
    +  You can highlight and run more than one line in a script with any of these methods

<br>
Notice how that read it straight into your console?

<br>
If you want to work with the data (which you do!), you have to assign a name to it.

<br>
`my_object <- read.csv("my_file.csv")`

<br>
__You can choose to name it whatever you want - but you should pick a name that makes sense!__

Trust us: your object name will help you follow along and will make it easier to remember if it makes sense.

```{r}
wqdat <- read.csv("gndpcwq2015.csv")
```

***

<br>

#### Step 3: See what your new object looks like

<br>
__Look at the head of the data__ (the first 6 rows)

<br>
Use the head() function:

```{r}
head(wqdat)
```

<br>
Notice, you did __not__ have to put _wqdat_ inside quotation marks. It is an object recognized by R. (You will also see it in the environment in the top right panel of RStudio.)

***

<br>
You can look at more rows if you want to. Let's look at the first 10 rows:

```{r}
head(wqdat, 10)
```

***

<br>

__Look at the end of the data__

<br>
Again, the default is 6 rows, but you can tell it to show you more.

```{r}
tail(wqdat)
```

***
<br>

__Look at summary statistics of the data__


This shows you a lot of useful information: min and max, quartiles, median and average, and the number of NAs.

<br>
__Note:__ NA is code for missing data.

<br>
_Hint:_ by looking at this output, you can see if there are major problems in your data.


```{r}
summary(wqdat)
```

***
<br>

__Look only at the column names__

This is especially helpful if you've been working on a script for a while but don't remember exactly how the name of your variable is showing up (capitalization, underscores, etc. - we name things differently sometimes in different files!)

```{r}
names(wqdat)
```


***

<br>
__Look at the structure of the dataset__

<br>
This shows you what _type_ of variable each column is: date, numeric, character, logical (true/false), etc.

```{r}
str(wqdat)
```


***

<br>

#### Step 4: Make a plot

<br>
Let's look at temperature over time. Basic plots are easy:

`plot(y ~ x, data = your data)`

<br>
In this specific case, it's:

`plot(Temp ~ DateTimeStamp, data=wqdat)`

<br>
But this throws errors! Let's look at just these two columns of data by using the $ selector:

```{r}
str(wqdat$Temp)
str(wqdat$DateTimeStamp)
```

***

<br>
__`DateTimeStamp`__ is seen as a factor. We need it to be seen as a date.

<br>
What we'll do is turn it into a __class__ called `POSIXct`, which is a good one to work with date-times. (There's also a class called `Date` if you're ever working with a simpler dataset - date format is a bit easier to work with.)

<br>
__This is something SWMPr does automatically__ as it reads in files, but we're not there yet in this course. So we'll do it by hand.

***

I'm going to make a new column, called `DateTime`. You could write over the original column if you want to. I'm hesitant to do that until I know I have something working.

<br>
We have to tell R what format the date/time in the current column is in, so that everything goes in the right place when it's turned into `POSIXct`. Notice the capitalization! (`%m` = month; `%M` = minutes)

<br>
To make a new column, we're doing the same thing we did when we created the data frame in the first place:

`Data_Frame$New_Column <- whatever we're doing to create the new column`


```{r}
wqdat$DateTime <- as.POSIXct(wqdat$DateTimeStamp, format="%m/%d/%Y %H:%M", tz="America/Regina")
```

<br>
Again, format is how you want your date to be read. `%m` is numerical month (01 = January), `%d` is numerical day (1-31), `%Y` is the numerical 4 digit year (ex. 2015), `%H` is hour, and `%M` is minute. This is how the SWMP data is formatted, so we’ll stick with it, but you could format months for example as `%b` (Jan, Feb, Mar, etc).

<br>
The timezone code matters - POSIXct will automatically adjust time stamps for Daylight Savings Time, and because we don't adjust our monitoring equipment, we definitely don't want R making that adjustment in the background.

Click here to find out which time zone is appropriate for you:
__need the link from Marcus__

***

<br>
We better check to make sure everything looks right before proceeding:

```{r}
head(wqdat)
str(wqdat)
```

***

<br>
Okay. Let's try this plot again:

```{r}
plot(Temp ~ DateTime, data=wqdat)
```


***

<br>
Let's change a few options - turn it into a line, and make that line red. Let's also give it some titles.

To figure out other plot options, type this into your console:

`?plot`


```{r}
plot(Temp ~ DateTime, data = wqdat,
     type = "l",
     col = "red",
     main = "Temperature at PC in 2015",
     xlab = "Date",
     ylab = "Temp (C)")
```


***

<br>
We can even change that y-axis to a more realistic scale:

```{r}
plot(Temp ~ DateTime, data = wqdat,
     type = "l",
     col = "red",
     main = "Temperature at PC in 2015",
     xlab = "Date",
     ylab = "Temp (C)",
     ylim = c(-5, 40))
```

***
<br>
We can also add that special degree symbol to make it publication-ready (_fun fact_: Kim learned this from Kari as they were planning this workshop):

```{r}
plot(Temp ~ DateTime, data = wqdat,
     type = "l",
     col = "red",
     main = "Temperature at PC in 2015",
     xlab = "Date",
     ylab = expression(paste("Temperature (",degree,"C)")),
     ylim = c(-5, 40))
```

***

<br>
__So we've already got several needs, and we're not even to the analysis stage yet:__

*  We need an easy way to get rid of bad data.
*  We need an easy way to turn date-time stamps into POSIXct.
*  We need an easy way to read in a lot of SWMP data and do each of the above.

<br>
__Enter the SWMPr package__

*  What is a package, generally, and why do I need them?
*  What are some popular packages?
    +  `ggplot2` for easy graphing
    +  `dplyr` and `tidyr` for data wrangling
    +  `wq` and `EnvStats` for statistics
    +  `openair` for wind roses
    +  `knitr` for generating nice documents from code and outputs
    +  `grDevices` for more color and font options
*  How do I get packages onto my computer?
    +  `install.packages("SWMPr")` for just one
    +  `install.packages(c("SWMPr", "tidyr", "ggplot2"))` for multiple
*  How do I use a package?
    +  load at the beginning of a session using either:
      +  `library(SWMPr)` in your script
      +  check `SWMPr` on the Packages tab in the bottom right panel of RStudio
    +  notice that you _do_ need quotation marks to install a package, but you _do not_ use them to load a package.
    
<br>
__Next, we'll use SWMPr to read in and QC some real data!__
    
<br>
<br>
.